Data Analysis

### 2025-07-12 docs/code/05_data_cleaning.Rmd
push to continue on laptop

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = TRUE)
```

@@

#### Setting up the space

```{r load_packages, message=FALSE, warning=FALSE}

@@

# write.csv(merged_data_extraction,here::here("data/03_data_cleaning/merged_data_extraction.csv"),row.names=FALSE)
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = TRUE)

if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman")
}

@@

@@

write.csv(merged_data_extraction,here::here("data/03_data_cleaning/merged_data_extraction.csv"),row.names=FALSE)
```

### 2025-07-12 docs/code/07_data_analysis.Rmd
push to continue on laptop

```{r}
# install.packages("pacman")
# load packages

pacman::p_load(rotl,
               ape,
               ggtree)

# We have 7 bird species in the dataset
bird_species <- unique(dataset_analysis$bird_species)

# extracting taxonomic information
taxa <- rotl::tnrs_match_names(names = bird_species)
taxa

# check approximate matches
taxa[taxa$approximate_match==TRUE & !(is.na(taxa$approximate_match)),]

# phylogenetic information for our species from the Open Tree of Life

tree <- rotl::tol_induced_subtree(ott_ids = taxa[["ott_id"]], label_format = "name")
ape::is.binary(tree)

# here are the species included in the tree
sort(tree$tip.label)

# removing the underscore "_" from the tree tip.label
tree$tip.label <- gsub("_"," ", tree$tip.label)
sort(tree$tip.label)

# compute branch lengths of tree
phylo_branch <- compute.brlen(tree, method = "Grafen", power = 1)

# check tree is ultrametric
is.ultrametric(phylo_branch) # TRUE

# matrix to be included in the models
bird_phylo <- vcv(phylo_branch, cor = T)
bird_phylo

plot(tree) #check: https://www.rdocumentation.org/packages/ape/versions/5.3/topics/plot.phylo
```

@@

## For lnRR (or simply load dataset as provided)
# dataset_lnRR<-dataset_analysis%>%
# filter(lnRR_sign != "NA" | lnRR_variance != "NA")

@@

## For SMDH
# selecting dataset which are not flagged
dataset_SMDH<-dataset_analysis%>%
   filter(proxy_decision == "include"| proxy_decision == "SMDH only"|proxy_decision =="contingency table") %>% filter(SMDH_sign != "NA")

@@

[**Robust variance estimation to guard against arbitary value of sampling correlation**]{.underline} - 'robust()' function from R package 'metafor' (Viechtbauer 2010) using 'CR2' (bias reduced linearization correction) from the R package 'clubSandwich' (Pustejovsky 2024) for small-sample size adjustment.

At the moment, this is not working..

```{r}
#MLMA_lnRR_RVE<- robust(MLMA_lnRR_VCV,
#                        cluster = paper_ID,
#                        clubSandwich = TRUE,
#                        verbose=T)

# # summary(MLMA_lnRR_RVE)
# I2_CI <- i2_ml(MLMA_lnRR_RVE) # Relative heterogeneity
# round(I2_CI,2) # absolute heterogeneity is reported in the summary of the model
```

::: callout-important
Sensitivity Analysis

We will also perform a sensitivity analysis using 0.3 and 0.7 correlation between the sampling variances from the same paper_ID and report it in the supplementary materials. If this sensitivity analysis shows large differences in conclusions for the intercept-only model, we will proceed to perform such sensitivity analysis for the meta-regressions.
:::

@@

```{r lnRR rho sensitivity analysis}
# rho_range <- c(0.3, 0.5, 0.7)
# 
# MLMAlnRR_VCV_range <- list() # repeatedly run the specified model with varying rho
# for (i in 1:length(rho_range))
# {VCV_range <- vcalc(vi = lnRR_variance,
#                   cluster = paper_ID,
#                   subgroup = group_ID,
#                   obs = Observation_ID,
#                   data = dataset_lnRR,
#                    rho = rho_range[i],
#                     ) # impute sampling variance covariance matrix with varying rho
# 
# MLMAlnRR_VCV_range[[i]] <- rma.mv(yi = lnRR_sign,
#                                   V = VCV_range, # sampling variance covariance matrix with varying values of rho.
#                                  random = list(~ 1 | paper_ID,
#                                   ~ 1 | Observation_ID,
#                                   ~ 1 | experiment_ID,
#                                   ~ 1 | group_ID,
#                                   ~ 1 | repeated_trait_ID),
#                                   method = "REML",
#                                   data = dataset_lnRR # run the model with varying rho
# )
# }
# 
# save(VCV_range, file = here::here("GNM/figures/VCV_range.Rdata"))
# save(MLMAlnRR_VCV_range, file = here::here("GNM/figures/MLMAlnRR_VCV_range.Rdata"))

@@

```{r Table for lnRR rho sensitivity analysis}

# t4 <- data.frame(rho  = rho_range,
#                  "overall effect"  = sapply(MLMAlnRR_VCV_range, function(x) coef(x)),
#                  "standard error" = sapply(MLMAlnRR_VCV_range, function(x) x$se),
#                  "p-value" = sapply(MLMAlnRR_VCV_range, function(x) x$pval),
#                  "Lower CI" = sapply(MLMAlnRR_VCV_range, function(x) x$ci.lb),
#                  "Upper CI" = sapply(MLMAlnRR_VCV_range, function(x) x$ci.ub),
#                  "Log-likehood" = sapply(MLMAlnRR_VCV_range, function(x) fitstats(x)[1,1]),
#                  "AIC" = sapply(MLMAlnRR_VCV_range, function(x) fitstats(x)[3,1]),
#                  "BIC" = sapply(MLMAlnRR_VCV_range, function(x) fitstats(x)[4,1]),
#                  "AICc" = sapply(MLMAlnRR_VCV_range, function(x) fitstats(x)[5,1]))
# 
# colnames(t4) <- c("Sampling correlation", "Overall effect (lnRR)", "Standard error", "p-value", "Lower CI", "Upper CI", "Log-likehood", "AIC", "BIC", "AICc")
# 
# t4 %>% dfround(4) %>% DT::datatable()
```

>>>>

```{r}
# # install.packages("pacman")
# # load packages
# 
# pacman::p_load(rotl,
#                ape,
#                ggtree)
# 
# 
# # We have 7 bird species in the dataset
# bird_species <- unique(dataset_analysis$bird_species)
# 
# # extracting taxonomic information
# taxa <- rotl::tnrs_match_names(names = bird_species)
# taxa
# 
# # check approximate matches
# taxa[taxa$approximate_match==TRUE & !(is.na(taxa$approximate_match)),]
# 
# # phylogenetic information for our species from the Open Tree of Life
# 
# tree <- rotl::tol_induced_subtree(ott_ids = taxa[["ott_id"]], label_format = "name")
# ape::is.binary(tree)
# 
# # here are the species included in the tree
# sort(tree$tip.label)
# 
# # removing the underscore "_" from the tree tip.label
# tree$tip.label <- gsub("_"," ", tree$tip.label)
# sort(tree$tip.label)
# 
# # compute branch lengths of tree
# phylo_branch <- compute.brlen(tree, method = "Grafen", power = 1)
# 
# # check tree is ultrametric
# is.ultrametric(phylo_branch) # TRUE
# 
# # matrix to be included in the models
# bird_phylo <- vcv(phylo_branch, cor = T)
# bird_phylo
# 
# plot(tree) #check: https://www.rdocumentation.org/packages/ape/versions/5.3/topics/plot.phylo
```

@@

## For lnRR (or simply load dataset as provided)
dataset_lnRR<-dataset_analysis%>%
filter(lnRR_sign != "NA" | lnRR_variance != "NA")

@@

## For SMDH
# selecting dataset which are not flagged
dataset_SMDH<-dataset_analysis%>%
   filter(proxy_decision != "sensitivity-analysis" | is.na(proxy_decision))

@@

::: callout-important
**Pre-registered**: We will also perform a sensitivity analysis using 0.3 and 0.7 correlation between the sampling variances from the same paper_ID and report it in the supplementary materials. If this sensitivity analysis shows large differences in conclusions for the intercept-only model, we will proceed to perform such sensitivity analysis for the meta-regressions.
:::

@@

```{r lnRR rho sensitivity analysis}

# rho_range <- c(0.2,0.3, 0.5, 0.7,0.8)
# 
# SA_lnRR_VCV_range <- list() # repeatedly run the specified model with varying rho
# for (i in 1:length(rho_range)){
#   VCV_range <- vcalc(vi = lnRR_variance,
#                      cluster = paper_ID,
#                      obs = Observation_ID,
#                      data = dataset_lnRR,
#                      rho = rho_range[i])
# 
#   lnRR_VCV_range[[i]] <- rma.mv(yi = lnRR_sign,
#                                 V = VCV_range,
#                                 random = list(~ 1 | paper_ID,
#                                               ~ 1 | Observation_ID,
#                                               ~ 1 | group_ID_coded,
#                                               ~ 1 | population_ID,
#                                               ~ 1 | bird_species),
#                                 method = "REML",
#                                 data = dataset_lnRR)
# }
# 
# 
# save(SA_lnRR_VCV_range, file = here::here("model/sensitivity_analysis/SA_lnRR_VCV_range.rds"))
# 
# 
# #------------------------------------------------------------------
# #-------------Now for SMDH the same SA with diff Rho---------------
# #------------------------------------------------------------------
# 
# SA_SMDH_VCV_range <- list()
# 
# for (i in 1:length(rho_range)){
#   VCV_range <- vcalc(vi = SMDH_variance,
#                      cluster = paper_ID,
#                      obs = Observation_ID,
#                      data = dataset_SMDH,
#                      rho = rho_range[i])
#   SA_SMDH_VCV_range[[i]] <- rma.mv(yi = SMDH_sign,
#                                 V = VCV_range,
#                                 random = list(~ 1 | paper_ID,
#                                               ~ 1 | Observation_ID,
#                                               ~ 1 | group_ID_coded,
#                                               ~ 1 | population_ID,
#                                               ~ 1 | bird_species),
#                                 method = "REML",
#                                 data = dataset_SMDH)
# }
# 
# 
# save(SA_SMDH_VCV_range, file = here::here("model/sensitivity_analysis/SA_SMDH_VCV_range.rds"))

@@
